---
title: "Luca_Epithelial_DEG"
output: html_document
date: "2023-09-21"
---


```{r}
library(Seurat)
library(dplyr)
library(tidyverse)

LucaExEpi <- readRDS("C:/Users/ferid/Downloads/lucaExEpi.rds")
```

# Create a new column with donor_id and ann_fine (condition) information to create the pseudobulk
```{r}
LucaExEpi$samples <- paste0(LucaExEpi$ann_fine, LucaExEpi$donor_id)
```

# Create the pseudobulk samples: Group the cells by donor and condition
```{r}
DefaultAssay(LucaExEpi)
cts <- AggregateExpression(LucaExEpi, 
                           group.by = "samples",  # Categories for grouping; 'ident' by default
                           assays = "RNA",        # Which assays to use. Default is all assays
                           slot = "counts",       # Slot(s) to use; if multiple slots are given, assumed to follow the order of 'assays' (if specified) or object's assays
                           return.seurat = FALSE) # Whether to return the data as a Seurat object. Default is FALSE

cts <- as.data.frame(cts$RNA)
```

# Generate sample level meta data (colData) with sample names
```{r}
colData <- data.frame(samples = colnames(cts))
```

# Add a condition column to colData based on sample names
```{r}
colData <- colData %>% 
  mutate(condition = ifelse(grepl("Cancer", samples), "Cancer", "Normal"))
```

# Add donor_id column to colData
```{r}
meta_data <- LucaExEpi@meta.data
donor_ids <- str_extract(colData$samples, paste(meta_data$donor_id, collapse = "|"))
colData$donor_id <- donor_ids
```

# Add rownames as samples
```{r}
rownames(colData) <- colData$samples
```

# Make sure all the column names of cts present in the row names of colData 
```{r}
all(colnames(cts) %in% rownames(colData))
```

# Make sure the column names of cts and the row names of colData are equal (in the same order)
```{r}
all(colnames(cts) == rownames(colData))
```

# Install DESeq2
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("DESeq2")

library(DESeq2)
```

# Create DESeq2 object with donor_id as a correction factor
```{r}
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = colData,
                              design = ~ donor_id + condition)
```

# Filter to remove all the genes that have lower than 100 reads 
```{r}
keep <- rowSums(counts(dds)) >= 100
dds <- dds[keep,]
```

# Error in estimateSizeFactorsForMatrix(counts(object), locfunc = locfunc, 
# every gene contains at least one zero, cannot compute log geometric means
# To avoid the error above;
# "Poscounts" considers only the counts that are greater than zero for each gene in each sample.
```{r}
dds <- estimateSizeFactors(dds, type = "poscounts")
```

# Reorder the levels of the "condition" factor variable to get the results of cancer compared to normal
# (Default first level is the first in alphabetical order)
```{r}
dds$condition <- relevel(dds$condition, ref = "Normal")
levels(dds$condition) # First level is the reference level
```

# Parallelization
```{r}
if (!requireNamespace("BiocParallel", quietly = TRUE))
  BiocManager::install("BiocParallel")

library(BiocParallel)
```

# Register the number of cores to use
```{r}
register(MulticoreParam(40))
```

# Enable parallelization for DESeq function and run
```{r}
dds <- DESeq(dds, parallel = TRUE)
saveRDS(dds, file = "dds.rds")
```

# Check the coefficients for the comparison
```{r}
resultsNames(dds)
```

# Generate results object. Alpha is the padj value treshold (alpha is 1 to keep all the genes) 
# Contrast is added to get the results of cancer compared to normal (if not specified before running DESeq)
```{r}
res <- results(dds, alpha = 0.99, contrast = c("condition", "Normal", "Cancer"))
res_df <- as.data.frame(res)
```

# Add a gene names column based on Ensembl gene ids 
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("org.Hs.eg.db")

library(org.Hs.eg.db)

ensembl_ids <- rownames(res_df)
gene_names <- mapIds(org.Hs.eg.db, keys = ensembl_ids, column = "SYMBOL", keytype = "ENSEMBL")
res_df$Gene_Name <- gene_names
```
